---
title: "Create Base Crime Data Frame"
author: "Jonathan Morken"
date: "March 19, 2016"
output: html_document
---

# Creating the base data file for later mapping, etc.
 
## Introduction
I will be combining data on Chicago crime patterns with other sources [geo-spatial polygons, Mass transit station locations, police stations locations, and demographics], in the hopes of finding robust evidence to suggest new station locations and/or policing patterns.  This file creates one large base CSV for use in later files.

### Sources
I am using crime data from the [Chicago Police Department](https://data.cityofchicago.org/), shape files and census data from the same source, and weather data from [Weather Underground](https://www.wunderground.com/history/airport/KMDW/2016/3/5/CustomHistory.html?req_city=Chicago%20Midway&req_state=IL&reqdb.zip=60499&reqdb.magic=11&reqdb.wmo=99999).  You can also pull this data from the data folder for this project on Github.  I also found a community-census tract mapping table courtesy of [Rob Paral](http://robparal.blogspot.com/2012/04/census-tracts-in-chicago-community.html).

Not all of the data is pulled in here, some is left for subsequent documents.  Note, I have intentionally created more data frames than necessary so any user can inspect each one for the differences.
```{r, message = FALSE, warning=FALSE}
#Load packages required for spatial statistics analysis
library(sp)
library(car)

#Packages for transforming data and calculating distances.
library(readr)
library(rgdal)
library(rgeos)
library(dplyr)
library(FNN)
library(magrittr)
library(lubridate) #dealing with time
library(classInt) #not using
```
 
### Read in the crime data  
```{r read_data}
setwd("D:/Users/Jonathan Morken/Documents/GitHub/Capstone Project/")
crime_df <- read.table("data/crimestxt.txt", sep = "\t", header = TRUE)
```

#### A quick look at the data structure shows 17 variables:
```{r view_structure}
str(crime_df)
```

#### We need to change the format of the following variables:
  - Make "Beat" a factor  
  - Make "Description" a factor  
  - Make "District" a factor  
  - Make "Community.Area" a factor  
  - Make "Date" POSIXct  
  - Split out time, day and month  
  - Subset the data to get rid of those records without lat/lons (done in a subsequent section).  

```{r format_variables}
crime_df$Beat <- as.factor(crime_df$Beat)
crime_df$Description <- as.factor(crime_df$Description)
crime_df$District <- as.factor(crime_df$District)
crime_df$Community.Area <- as.factor(crime_df$Community.Area)
crime_df$Date <- parse_date_time(crime_df$Date, orders="mdy hm", tz = "America/Chicago")
crime_df <- mutate(crime_df,Month = month(Date))
crime_df <- mutate(crime_df,DayinMonth = mday(Date))
crime_df <- mutate(crime_df,DayofWeek = wday(Date, label = TRUE, abbr = TRUE))
crime_df <- mutate(crime_df,Hour = hour(Date))
```

### Pull in and format other data including weather, census, crime categories, hour categories, transit locations, and police department locations.
Feel free to explore the resulting data frames on your own.
```{r add_additional_data}

#Weather Data (Recode "trace" precipitation = 0, add year month and day for joining data)
weather_df <- read.csv("data/WeatherReduced.csv") 
weather_df$Date <- parse_date_time(weather_df$Date, orders="mdy hm")
weather_df$PrecipitationIn <- recode(weather_df$PrecipitationIn, "T = 0")
weather_df$PrecipitationIn <- as.integer(weather_df$PrecipitationIn)
weather_df <- mutate(weather_df,Year = year(Date))
weather_df <- mutate(weather_df,Month = month(Date))
weather_df <- mutate(weather_df,DayinMonth = mday(Date))

#Census and Population Data (Limit columns, classify community area as factor, join population and general census data.)
census_df <- read.csv("data/CensusData.csv", sep = "\t", header = TRUE)
census_df$Community.Area <- as.factor(census_df$Community.Area)
population_df <- read.csv("data/PopulationByCommunity.csv", sep = ",", header = TRUE)
population_df$Community.Area <- as.factor(population_df$Community.Area)
census_df <- left_join(census_df,population_df, by = "Community.Area")

#Crime Category Mapping
Crime_Class_Map <- read.table("data/CrimeClassMapping.txt", sep = "\t", header = TRUE)

#Hour Categories Mapping
Hour_Categories <- read.csv("data/HourCategories.csv", sep = ",", header = TRUE)

#Pull in Mass Transit and Police Department Locations and create separate data frames for easy inspection later.
Mass_Transit <- read.csv("Data/CTAListofLStops.csv", sep = ",", header = TRUE)
Mass_Transit <- Mass_Transit[,c(1,2,3,15,16)]
Transit_Coord <- Mass_Transit

Police_Dps <- read.csv("Data/PoliceStations.csv", sep = ",", header = TRUE)
Police_Dps <- Police_Dps[,c(1,6,7)]
Police_Coord <- Police_Dps
```

#### Join census, weather, crime categories and hour categories to crime data
```{r Join_to_Crime Data, warning=FALSE}

#Join to crime_df
crime_df <- inner_join(crime_df,Crime_Class_Map,by = c("Primary.Type","Description"))
crime_df <- left_join(crime_df,census_df,by = "Community.Area")
crime_df <- left_join(crime_df,weather_df,by = c("Year","Month","DayinMonth"))
crime_df$Community.Area <- as.factor(crime_df$Community.Area)
crime_df$Year <- as.factor(crime_df$Year)
crime_df <- left_join(crime_df,Hour_Categories, by = c("Hour"))
```

### Calculating Proximity Variables

#### Make coordinates planar for determining closest PD and transit station 
I.e. We will use these data frames for calculating which PD/Transit station is closest to each crime and which PD is closest to each transit station.
```{r Make_Planar}

#Mass Transit
coordinates(Transit_Coord) <- ~ Longitude+Latitude
proj4string(Transit_Coord) <- CRS("+init=epsg:4326") #Latitude/Longitude
data.Transit <- spTransform(Transit_Coord, CRS("+init=epsg:2790")) # set for IL plane

#Police Departments
coordinates(Police_Coord) <- ~ Longitude+Latitude
proj4string(Police_Coord) <- CRS("+init=epsg:4326") #Latitude/Longitude
data.PD <- spTransform(Police_Coord, CRS("+init=epsg:2790")) #set for IL plane

#Crimes
crime_df_no_NAs <- subset(crime_df, !is.na(Longitude)) #Remove NAs from Crime data
Crime_Planar <- crime_df_no_NAs[, c(1,15,16)]          #Shrink crime data frame for the sake of memory.  
coordinates(Crime_Planar) <- ~ Longitude+Latitude
proj4string(Crime_Planar) <- CRS("+init=epsg:4326") #Latitude/Longitude
data.CrimePlanar <- spTransform(Crime_Planar, CRS("+init=epsg:2790")) #set for IL plane
```

#### Get closest PD and Transit location for each crime then join to Crime DataFrame
```{r Ascertain_Closest_Stations}

#Police Station Closest to Each Crime
d <- get.knnx(coordinates(Police_Coord), coordinates(Crime_Planar), k=1)
# str(d)

#Transit Station Closest to Each Crime
d2 <- get.knnx(coordinates(Transit_Coord), coordinates(Crime_Planar), k=1)
# str(d2)

#Police Department closest to Each Transit Station
d3 <- get.knnx(coordinates(Police_Coord), coordinates(Transit_Coord), k=1)
# str(d3)

#Convert results to data frames; note that the distance results are in degrees and not quite correct when transformed into kilometers.  As such, we will recalculate these distances via a formula below.
d_df <- as.data.frame(d)
d_df <- rename(d_df, PD.index = nn.index)
d_df <- rename(d_df, PD.dist = nn.dist)
d2_df <- as.data.frame(d2)
d2_df <- rename(d2_df, Transit.index = nn.index)
d2_df <- rename(d2_df, Transit.dist = nn.dist)
d3_df <- as.data.frame(d3)
d3_df <- rename(d3_df, PD.index = nn.index)
d3_df <- rename(d3_df, PD.dist = nn.dist)

#Bind d and d2 to original data frame. 
crime_df_no_NAs <- bind_cols(crime_df_no_NAs,d_df,d2_df)
crime_df_no_NAs <- as.data.frame(crime_df_no_NAs)

#Change PD and Transit data frames to include row number
Police_Dps$PD.index <- seq.int(nrow(Police_Dps))
Mass_Transit$Transit.index <- seq.int(nrow(Mass_Transit))

#Join closest PD and transit coordinates to crime data for correctly calculating distance.
crime_df_no_NAs <- left_join(crime_df_no_NAs,Police_Dps, by = "PD.index")
crime_df_no_NAs <- left_join(crime_df_no_NAs,Mass_Transit, by = "Transit.index")
crime_df_no_NAs <- rename(crime_df_no_NAs, Longitude.PD = Longitude.y)
crime_df_no_NAs <- rename(crime_df_no_NAs, Latitude.PD = Latitude.y)
crime_df_no_NAs <- rename(crime_df_no_NAs, Longitude.Transit = Longitude)
crime_df_no_NAs <- rename(crime_df_no_NAs, Latitude.Transit = Latitude)
crime_df_no_NAs <- rename(crime_df_no_NAs, Longitude.Crime = Longitude.x)
crime_df_no_NAs <- rename(crime_df_no_NAs, Latitude.Crime = Latitude.x)
crime_df_no_NAs <- rename(crime_df_no_NAs, Closest.District = DISTRICT)
crime_df_no_NAs <- rename(crime_df_no_NAs, Community.Name = COMMUNITY.AREA.NAME.x)
crime_df_no_NAs <- rename(crime_df_no_NAs, Station.Name = STATION_NAME)

#Bind d3 to Mass_Transit data frame
Mass_Transit <- bind_cols(Mass_Transit,d3_df)
Mass_Transit <- as.data.frame(Mass_Transit)

#Join closest PD from d3 to transit data for correctly caluclating distance.
Mass_Transit <- left_join(Mass_Transit,Police_Dps, by = "PD.index")
Mass_Transit <- rename(Mass_Transit, Longitude.PD = Longitude.y)
Mass_Transit <- rename(Mass_Transit, Latitude.PD = Latitude.y)
Mass_Transit <- rename(Mass_Transit, Longitude = Longitude.x)
Mass_Transit <- rename(Mass_Transit, Latitude = Latitude.x)
```

#### Create formula to better approximate Euclidian distance between two points in kilometers
```{r Formula_for_correct_Euclidian_distance}
#Create formula to calculate distance between two points in kilometers (this calculation can also be found in the "fossil" package.  Note 6378.145 is the avg of the earth's radius.
earth.dist <- function (long1, lat1, long2, lat2)
{
rad <- pi/180
a1 <- lat1 * rad
a2 <- long1 * rad
b1 <- lat2 * rad
b2 <- long2 * rad
dlon <- b2 - a2
dlat <- b1 - a1
a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
c <- 2 * atan2(sqrt(a), sqrt(1 - a))
R <- 6378.145
d <- R * c
return(d)
}
```

#### Add correct distances to crime data and mass transit data frame
Note, you could replace PD.dist and Transit.dist instead of creating a new variable, but i am leaving them along for the viewer's sake. 
```{r Calculate_correct_distances}
crime_df_no_NAs$PD.Dist.km <- earth.dist(crime_df_no_NAs$Longitude.Crime,crime_df_no_NAs$Latitude.Crime,crime_df_no_NAs$Longitude.PD,crime_df_no_NAs$Latitude.PD)

crime_df_no_NAs$Transit.Dist.km <- earth.dist(crime_df_no_NAs$Longitude.Crime,crime_df_no_NAs$Latitude.Crime,crime_df_no_NAs$Longitude.Transit,crime_df_no_NAs$Latitude.Transit)

Mass_Transit$PD.Dist.km <- earth.dist(Mass_Transit$Longitude,Mass_Transit$Latitude,Mass_Transit$Longitude.PD,Mass_Transit$Latitude.PD)
```

### Write final mass data files
Again, I would normally recommend replacing data frames and variables along the way, but i did not do that on this exercise for the sake of the viewer tracking differences.  We will cut out excess columns in subsequent analysis files.
```{r Write_Final_Data_Files}
#Write joined files [crime superset file and Mass transit file with distance to closest PD]:
#write.csv(crime_df_no_NAs, file = "Data/JoinedDataWithDist.csv", row.names = FALSE)
#write.csv(Mass_Transit, file = "Data/CTAListofLStopsWithPDDistance.csv", row.names = FALSE)

```

## See the next file for analysis and plots.